Appendix: Trends in age-adjusted mortality rates for blacks and whites by cause and sex, 1969-2013, United States

### 2017-03-29 Manuscript/Manuscript_COD_Trends.Rmd
new manuscript

---
title: Exploring how changes in cause of death have narrowed the black-white life
  expectancy gap in the United States.
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

@@

The overall difference in life expectancy between black and whites in the US has generally decreased over time. States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York experiencing greater reductions to the gap compared to states such as California and Texas. The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in narrower life expectancy gaps in some states and  wider gaps in other states.

@@

Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes). We excluded 10 states with the smallest black populations, but included the District of Columbia. A Bayesian autoregressive model was used to model smoothed mortality rates over time within each stratum, and a truncated Poisson model was used to impute death counts that were suppressed due to small cell sizes. We calculated life expectancy for blacks and whites within each stratum using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method.

[Add findings]

**Conclusions:**
[Add conclusion]

###Introduction
The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this figure are differences by race. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [Harper 2007]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014]. This study seeks to better understand these state-level differences, by examining how changes in the cause of death over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013. Common reasons for the decreasing life expectancy gap can thereby be identified, and anomalous states can be examined. Ideally, one would like to identify policies or social conditions in highly successful states that help to explain their unusual performance.

@@

**The data**

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.

@@

**Statistical Methods**

The data structure produced two statistical challenges that were addressed by our quantitative approach. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9). These models were run separately for every combination of state, race, and sex, and produced death counts that were smoothed over time for every cause of death within each age group.

@@

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref]. First, the model partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks. For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites. This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination. For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old. We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference, for each state and for each sex.

In order to understand whether states fared similarly over time, we divided the time frame into three eras which roughly encapsulate an initial period of increases in life expectancy for both races and sexes (1969-1983), an intermediate period of decreases and stagnating life expectancy for blacks but not for whites (1983-1993) and a final period of increasing life expectancy for both races (1993-2013). While these eras represent different lengths of time, they best capture three distinct periods of changes to the difference in life expectancy: an intial narrowing of the gap, followed by a widening of the gap that diminished previous progress, and the most recent twenty years of sustained narrowing (**S1 Fig** and **S2 Fig**).

For each era, we calculated the degree that the life expectancy gap has narrowed or widened and decomposed this *change* to the *difference* in the life expectancy between blacks and whites according to cause of death. To do so, we simply subtracted the contribution (in years) that each cause of death category had made to the difference in life expectancy at the beginning of the era from that made at the beginning of the era. An example of the calculation is found in **S1 Text**.

*ktm to read over*: The bayesian model was fit using two markov chains of length 10,000. One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation). These 1,000 iterations approximate the shape of the posterior distribution, and we used the mean as the estimate of the log mortality rate, and the 2.5th and 97.5th percentiles as the lower and upper bound of the 95% credible interval (CI).

All data analyses were performed using R version 3.3.1. The Bayesian smoothing was performed using `rjags` (version 4-6) and `R2jags` (version 0.5-7), and the life expectancy and decomposition analysis were performed using user-written functions found in the Supplement.

@@

```{r initialize, echo=FALSE, warning=F, message=F}
library(shiny)
library(plotly)
library(viridis)
library(scales)
library(shinythemes)
library(dplyr)
library(purrr)
library(stringr)
library(DT)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")
```

```{r read_data, echo=F}
age_cod_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_results.csv")
age_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_results.csv")
cod_decomp_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_decomp_results.csv")
cod_change_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_change_results.csv")
BlackWhite_results <- read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_results.csv")

age_cod_results$COD <- factor(age_cod_results$COD, levels(age_cod_results$COD)[c(3, 2, 4, 6, 5, 1)])
cod_decomp_results$COD <- factor(cod_decomp_results$COD, levels(cod_decomp_results$COD)[c(3, 2, 4, 6, 5, 1)])
age_decomp_results$age <- factor(age_decomp_results$age, levels = levels(age_decomp_results$age)[c(1, 2, 11, 3:10, 12:19)])
age_cod_results$age <- factor(age_cod_results$age, levels = levels(age_cod_results$age)[c(1, 2, 11, 3:10, 12:19)])

cod_change_results$Cause.of.death <- factor(cod_change_results$Cause.of.death, levels(cod_change_results$Cause.of.death)[c(3, 2, 4, 6, 5, 1)])
```

```{r prep_era_figure, echo = F}
year1 <- 1969
year2 <- 1983
year3 <- 1993
year4 <- 2013

bw_dat <- BlackWhite_results %>% 
  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap_mean) == F) %>%
  select(state, year, LE_wbgap_mean, sex) %>%
  arrange(year)

bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap_mean) 
names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  
bw_dat.males <- bw_dat %>% filter(sex == "Male") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )

bw_dat.females <- bw_dat %>% filter(sex == "Female") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )
```

```{r more_figure_prep, echo = F}
widths <- cod_change_results %>% mutate(tot_Cx_era1_CI_width = tot_Cx_era1_diff_ucl - tot_Cx_era1_diff_lcl,
                                        tot_Cx_era2_CI_width = tot_Cx_era2_diff_ucl - tot_Cx_era2_diff_lcl,
                                        tot_Cx_era3_CI_width = tot_Cx_era3_diff_ucl - tot_Cx_era3_diff_lcl) %>%
  filter(Cause.of.death == "Injuries") %>%
  select(state, sex, tot_Cx_era1_CI_width, tot_Cx_era2_CI_width, tot_Cx_era3_CI_width)

widths <- tidyr::gather(widths, tot_Cx_era1_CI_width, tot_Cx_era2_CI_width, tot_Cx_era3_CI_width, key = "era", value = "tot_Cx_CI_width")
ncols <- str_split_fixed(as.character(widths$era), "_", 5)
widths$era <- as.factor(ncols[ , 3])

cod.df <- cod_change_results %>% select(state, sex, Cause.of.death, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean)
cod.df <- tidyr::gather(cod.df, Cx_era1_diff_mean, Cx_era2_diff_mean, Cx_era3_diff_mean, key = "era", value = "Cx_diff_mean")
ncols <- str_split_fixed(as.character(cod.df$era), "_", 4)
cod.df$era <- as.factor(ncols[ , 2])

cod.df2 <- cod_change_results %>% select(state, sex, Cause.of.death, Cx_p_era1_diff_mean, Cx_p_era2_diff_mean, Cx_p_era3_diff_mean)
cod.df2 <- tidyr::gather(cod.df2, Cx_p_era1_diff_mean, Cx_p_era2_diff_mean, Cx_p_era3_diff_mean, key = "era", value = "Cx_p_diff_mean")
ncols <- str_split_fixed(as.character(cod.df2$era), "_", 5)
cod.df2$era <- as.factor(ncols[ , 3])

cod.df <- merge(cod.df, cod.df2, by = c("state", "sex", "Cause.of.death", "era"))

cod.df <- merge(cod.df, widths, by = c("state", "sex", "era"))

cod.df$narrowed_gap <- T
cod.df$narrowed_gap[cod.df$Cx_diff_mean < 0 ] <- F

cod.df <- make_pretty_decomp_plot(decomp.table = cod.df, strat.var.1 = "state",
                                  strat.var.2 = "sex", strat.var.3 = "era", sign.var = "narrowed_gap",
                                  decomp.var = "Cx_diff_mean",
                                  decomp.var.prop = "Cx_p_diff_mean",
                                  partition.bar.var = "Cause.of.death",
                                  type.of.data = "results")

cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
                                           c("state", "Census_Region", "Census_Division", "stabbrs")], by = c("state"))
cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)

cod.df.males$tot_Cx_CI_width2 <- -(cod.df.males$tot_Cx_CI_width - min(cod.df.males$tot_Cx_CI_width))/(max(cod.df.males$tot_Cx_CI_width) - min(cod.df.males$tot_Cx_CI_width)) + 1
cod.df.males$width.factor <- as.factor(cut(cod.df.males$tot_Cx_CI_width2, breaks = c(-0.01, 0.25, 0.50, 0.75, 1)))

```

*Changes to the difference in life expectancy in US males, 1969-1983*

**Fig 1** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs. their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines along the x-axis. This gap is decomposed by cause of death, where causes to the left of the dashed line contributed to narrowing the gap and causes to the right contributed to widening the gap. The length of these coloured bars represent the number of years attributable to the cause along the x-axis.

For example, in the **Fig 1**, New York's gap changed from 8.8 years in 1969 to 6.3 in 1983. Causes to the left of the dashed line -- All other causes (pink), injuries (blue), non-communicable diseases (teal), and communicable diseases (green) -- each contributed to narrowing the gap, while cardiovascular disease (red) and cancer (yellow) contributed to widening the gap. Since the contribution in years of the leftward causes is larger than the contribution of the rightward causes, the net change is a smaller life expectancy gap in 1983, where all other causes (the longest bar) made the largest contribution.

```{r, echo = F}
sum1 <- BlackWhite_results %>% filter(year == 1969 & sex == "Male") %>% 
  arrange(LE_wbgap_mean) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)

diff <- bw_dat.males %>% arrange(era1.diff)

for(i in 1:5) {
  j <- i-1

  max.value <- max(cod.df$Cx_diff_mean[cod.df$state == diff[dim(diff)[1]-j, ]$state & 
                                            cod.df$era == "era1" &
                                            cod.df$sex == "Male" ])

  max.cause <- cod.df$Cause.of.death[cod.df$state == diff[dim(diff)[1]-j, ]$state &
                    cod.df$sex == "Male" &
                    cod.df$era == "era1" & 
                    cod.df$Cx_diff_mean == max.value]

  min.value <- min(cod.df$Cx_diff_mean[cod.df$state == diff[i, ]$state & 
                                            cod.df$era == "era1" &
                                            cod.df$sex == "Male" ])

  min.cause <- cod.df$Cause.of.death[cod.df$state == diff[i, ]$state &
                    cod.df$sex == "Male" &
                    cod.df$era == "era1" & 
                    cod.df$Cx_diff_mean == min.value]

  assign(paste0("max", i), max.value)
  assign(paste0("max.cause", i), max.cause)
  assign(paste0("min", i), min.value)
  assign(paste0("min.cause", i), min.cause)
}
```

```{r, echo =F}
sum2 <- BlackWhite_results %>% filter(year == 1983 & sex == "Male") %>% 
  arrange(LE_wbgap_mean) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)

  min.value.DC <- min(cod.df$Cx_diff_mean[cod.df$state == "Washington DC" & 
                                            cod.df$era == "era2" &
                                            cod.df$sex == "Male" ])

  min.cause.DC <- cod.df$Cause.of.death[cod.df$state == "Washington DC" &
                    cod.df$sex == "Male" &
                    cod.df$era == "era2" & 
                    cod.df$Cx_diff_mean == min.value.DC]
```

```{r, echo =F}
sum3 <- BlackWhite_results %>% filter(year == 1993 & sex == "Male") %>% 
  arrange(LE_wbgap_mean) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)

sum4 <- BlackWhite_results %>% filter(year == 2013 & sex == "Male") %>% 
  arrange(LE_wbgap_mean) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap_mean, LE_wbgap_ucl)
```

In 1969, the life expectancy gap ranged from `r round(sum1[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap_mean, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `. Injury was the largest contributor towards narrowing the gap during this time period, as both race groups experienced decreased rates of injury-related mortality, but blacks had larger rates decreases, especially in states with relatively higher rates in 1969 (**S3-S14 Figs**). A few states, including Minnesota and Nebraska, had relatively low rates of injuries among blacks in 1969 and smaller gaps. The residual category also contributed to narrower gaps in most states, but generally made a smaller contribution than injuries. In New York, the residual category made a large contribution to narrowing the gap, where much of this decrease was likely due to decreased infant mortality among blacks (**S15 Fig**). Communicable diseases contributed modestly to narrowing the gap, except in Arizona and Minnesota. Generally, cardiovascular disease contributed towards increasing the gap, and reflects slower declines for blacks in cardiovascular mortality than for whites. Cancer contributed towards increasing the gap in most states, and largely reflects increased cancer-related mortality for both races during this period, with blacks undergoing steeper changes. In most states, non-communicable diseases contributed a small amount towards increasing the gap, although it decreased the gap among four states, most notably in New York.

For `r dim(bw_dat.males[bw_dat.males$era1.diff > 0,])[1]` of the `r dim(bw_dat.males)[1]` states, the gap decreased between 1969 and 1983, with the largest decreases in `r diff[dim(diff)[1], ]$state` (decrease in the gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)` years, `r round(max1,1)` years, of which is due to `r max.cause1`), `r diff[dim(diff)[1]-1, ]$state` (change of `r round(diff[dim(diff)[1]-1, ]$era1.diff, 1)` years, `r round(max2,1)` due to `r max.cause2`), `r diff[dim(diff)[1]-2, ]$state` (change of `r round(diff[dim(diff)[1]-2, ]$era1.diff, 1)` years, `r round(max3,1)` due to `r max.cause3`), `r diff[dim(diff)[1]-3, ]$state` (change of `r round(diff[dim(diff)[1]-3, ]$era1.diff, 1)` years, `r round(max4,1)` due to `r max.cause4`), and `r diff[dim(diff)[1]-4, ]$state` (change of `r round(diff[dim(diff)[1]-4, ]$era1.diff, 1)` years, `r round(max5,1)` due to `r max.cause5`).

The gap increased for `r dim(bw_dat.males[bw_dat.males$era1.diff < 0,])[1]` states. Most notably in `r diff$state[1]` (increase in gap of `r round(abs(diff$era1.diff[1]), 1)` years, `r abs(round(min1,1))` due to `r min.cause1`), `r diff$state[2]` (increase in gap of `r round(abs(diff$era1.diff[2]), 1)` years, `r abs(round(min2,1))` due to `r min.cause2`), `r diff$state[3]` (increase in gap of `r round(abs(diff$era1.diff[3]), 1)` years, `r abs(round(min3,1))` due to `r min.cause3`), `r diff$state[4]` (increase in gap of `r round(abs(diff$era1.diff[4]), 1)` years, `r abs(round(min4,1))` due to `r min.cause4`). For these states, injuries do not contribute as much to narrowing the gap as in other states (and in Minnesota, injuries widened the gap), so these disparity increases appear to be due to a combination of changes to cardiovascular-related mortality making blacks relatively worse off and less changes to injuries than experienced by other states.

During the second time period, from 1983 to 1993, the life expectancy gap for males widened in `r dim(bw_dat.males[bw_dat.males$era2.diff < 0, ])[1]` of the `r dim(bw_dat.males)[1]` states (**Fig 2**). For states such as New York, Maryland, New Jersey, and Connecticut, the biggest driver of the widening gap was communicable disease, reflecting the heightened burden from HIV/AIDS in black males during the height of the AIDS epidemic. While both blacks and whites experienced a dramatic increase in mortality related to communicable diseases, blacks' rates were even higher. Injuries also contributed to the gap widening, most notably in Wisconsin, Maryland, and Louisiana. For many states, injury-related mortality increased for blacks but not for whites during this time period. Compared to the previous period, these two causes of death led to widwer rather than narrowed gaps. Cancer and cardiovascular disease contributed towards wider gaps, although the magnitude of the contribution of cardiovascular disease is quite hetergenous by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions. The residual category contributed to a narrower gap, especially in New York, Pennsylvania, Connecticut, Mississippi, and Florida [note: This wasn't related to infant mortality improvements among blacks in NY, PA, but partially related to IMR improvements in CT, MS, and FL]. Most notably, the District of Columbia had the highest gap to begin and underwent the largest increase to their gap, which increased from 10.5 years to 15.9 years over this 10-year period, reflecting an increase in the gap of `r abs(round(bw_dat.males$era2.diff[bw_dat.males$sex == "Male" & bw_dat.males$state == "Washington DC"], 1))` years, of which `r round(min.value.DC, 1)` was due to `r min.cause.DC`.

The final time period saw decreasing gaps in all states, and an increasing gap in Washington DC (**Fig 3**). In 1993, the gap ranged from `r round(sum3[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum3[dim(sum3)[1], ]$state `. By 2013, the gap ranged from `r round(sum4[1, ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum4[1, ]$LE_wbgap_lcl, 1)`, `r round(sum4[1, ]$LE_wbgap_ucl, 1)`] in `r sum4[1, ]$state ` to `r round(sum4[dim(sum4)[1], ]$LE_wbgap_mean, 1)` years [95% CI: `r round(sum4[dim(sum4)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum4[dim(sum4)[1], ]$LE_wbgap_ucl, 1)`]  in `r sum4[dim(sum4)[1], ]$state `. During this time period all of the causes contributed to narrowing the gap for most states. In Wisconsin and the District of Columbia, changes to CVD mortality counteracted the gap narrowing (reflecting slower decreases to cardiovascular mortality among blacks compared with whites), while in Arizona changes to communicable disease counteracted the narrowing.

**Fig 1: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**
```{r figure1, fig.align = "center", echo=F, fig.height=9, fig.width=7}
y.labels <- levels(factor(cod.df.males$state.gdiff.order.1))
y.num.breaks <- length(levels(factor(cod.df.males$state.gdiff.order.1)))

males.era1 <- ggplotly(
  ggplot(subset(cod.df.males, era == "era1"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = Cause.of.death, alpha = width.factor), 
              color = "white") +
    geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
        geom_segment(data = subset(cod.df.males, Cause.of.death == "Cancers" & era == "era1"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
    theme_minimal() + 
    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era1
```

**Fig 2: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**

```{r figure2, echo = F, fig.height=9, fig.width=10}
# - The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.

y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
y.num.breaks.2 <- length(levels(factor(cod.df.males$state.gdiff.order.2)))

males.era2 <- ggplotly(
  ggplot(subset(cod.df.males, era == "era2"), aes(y = state.gdiff.num.2, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                  ymax = state.gdiff.num.2 + 0.45, 
                  xmax = new.finish.2, fill = Cause.of.death, alpha = width.factor), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
    theme_minimal() + 
    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2), minor_breaks = NULL) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era2 
```

**Fig 3: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1993 and 2013 by US state**

```{r figure3, echo = F, fig.height = 9, fig.width = 11}
y.labels.3 <- levels(factor(cod.df.males$state.gdiff.order.3))
y.num.breaks.3 <- length(levels(factor(cod.df.males$state.gdiff.order.3)))

>>>>

---
title: 
output:
  html_document: default
  word_document: default
always_allow_html: yes
---

@@

@@

**Conclusions:**

###Introduction
The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this figure are differences by race. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.

During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [@harper2007trends]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [@harper2014trends].

The over-arching objective of this study was to describe the general pattern of each cause of death's contribution to the black-white life expectancy gap over forty-five years, and identify divergent states and differences between the sexes. To do so, we estimated state-level trends in the contribution of six major causes of death to the black-white life expectancy gap for males and females. Common reasons for the decreasing life expectancy gap can thereby be identified, and anomalous states can be examined. Ideally, this study may facilitate the identification of one policies or social conditions in highly successful states, and suggest broad targets for interventions to further reduce the gap in less successful states.

@@

**The data**
We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia between 1969 and 2013. The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data were derived by the NCHS using intercensal bridged-race population estimates. This study used data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**S1 Table**). When the number of deaths for a population stratum is between 1 and 9, this data point is suppressed by the software and produces a missing death count for that cell However, death counts of exactly 0 or greater than 9 are known precisely.

@@

**Statistical Methods**
The data structure produced two statistical challenges that were addressed by our quantitative approach. First, our methodology needed to impute the suppressed death counts (for counts between 1 and 9). Second, several states had small black populations which produced sparse counts in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modeled autoregressively, where the log mortality rate in the current year was modeled as a normal variable centered at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the suppressed death counts, by specifying that the suppressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9).

These models were run separately for every combination of state, race, and sex. They were fit using two markov chains of length 10,000. One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation). Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.

@@

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref]. First, the model partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks. For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites. This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination. For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old. We summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference, for each state and for each sex. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 45 state-level trends and their credible intervals.

To describe the general pattern in these trend lines, we regressed each state's mean contribution against a linear spline for calendar time that was estimated with five degrees of freedom. This model also contained fixed effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year. We then calculated residuals for each year by substracting the meta-estimated contribution from the smoothed-observed estimated contribution and summed these residuals over time for each state to rank states according to how well the model predicted their pattern. Plots of these residuals over time were used to identify divergent states.

**Need to add information on the trends in mortality**

@@

###Bibliography
